Beyond Mechanistic Models: The case for machine learning in ecological management

### 2013-06-06 inst/doc/beyond-mechanistic-models.tex
manuscript helper files

% Beyond Mechanistic Models: The case for machine learning in ecological management

Abstract
============

<!-- Run R code for analysis, to be called by figures -->
```{r, echo=FALSE, cache=FALSE, include=FALSE}
options(xtable.print.comment=FALSE)
options(xtable.type = 'latex', table.placement="H")
opts_chunk$set(warning=FALSE, message=FALSE, comment=NA,
               tidy=FALSE, echo=FALSE)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(dev = 'Cairo_pdf', fig.width=5.5, fig.height=4,
 cache.path = 'cache-pdf/', cache=TRUE, external=TRUE)
read_chunk("nonparametric-bayes.R")
```
```{r knit-settings, echo=FALSE, include=FALSE, cache=FALSE}
```
```{r libraries, echo=FALSE, include=FALSE}
```
```{r plotting-options, echo=FALSE, include=FALSE}
```
```{r stateeq}
```
```{r sdp-pars, dependson="stateeq"}
```
```{r obs, dependson="sdp-pars"}
```
```{r mle, dependson="obs"}
```
```{r gp-priors}
```
```{r gp, dependson=c("obs", "gp-priors")}
```
```{r opt, dependson=c("gp", "mle")}
```
```{r sim, dependson="opt"}
```

Introduction
===============================================================

@@

While effective machine learning algorithms have speckled the literature, 
most theoretical ecologists in particular have been wary of these 'black-box'
approaches in favor of simple, mechanistic models.  While a blind application
of machine learning approaches would indeed be dangerous grounds to stake 
either ecological understanding or management practice, such concerns are not
grounds to reject such methods out of hand.  Rather, further attention is 
warranted to identify the circumstances in which machine-learning approaches
will most likely outperform existing more mechanistic approaches, or help
crack difficult management questions where ecological or dynamical complexity
has caused traditional approaches to founder.

@@

Mechanistic models have long been the gold standard of theoretical
modeling in ecology (e.g. see @Geritz2012 or @Cuddington2013).  Only by
understanding the processes involved can we make reliable long term
predictions and build an knowledge of cause and effect that guides the
hypotheses we make, the data we collect, and the management decisions
we make.  Process-based models, whether expressed in the language of
mathematics or English, identify the connection between mosquitoes and
the spread of malaria, or greenhouses gases and climate change, guiding
our approach to understand and manage these threats. Despite this central
importance, we argue here that ecologists would do well to give greater
attention to the role non-mechanistic models can play in ecological
management and decision making.  The value of these approaches is
greatest in a context where decisions are made over short time horizons,
and updated as new data becomes available.

## The alternative to process-based models: pattern-based modeling

Before we proceed further it would be useful to define our terms and lay
some greater context for the discussion. Throughout, we will distinguish
"mechanistic" or "process-based" models from models that are merely
"correlative" or "pattern-based". Historically pattern-based modeling
meant regression (usually linear regression) and resided in methodology
rooted in statistics departments and was the primary focus of empirical
ecologists, while mechanistic modeling meant dynamical systems (ODEs,
PDEs, SDEs and their discrete kin), residing in methodology from
the mathematics department and was the primary focus of theoretical
ecologists.

We believe these historical divides continue to color much of the
literature today, with theorists more skeptical than empiricists of
statistical methodology and vice versa.  Meanwhile, the ground underneath
has shifted.  Originally pattern-matching approaches could be critiqued
on the grounds of their simplicity (not everything is linear) while
mechanistic approaches could be critiqued on their tenuous connections
to data -- model parameter values such as death rates, birth rates,
etc. would be estimated in advance and then stuck into the model, rather
than estimated in the context of the model itself.  Today both of these
critiques are outdated. Computational power and hierarchical statistical
methods (particularly approximate likelihood or simulation techniques such
as particle filters and approximate Bayesian computing) have brought the
inference richer dynamical systems into their fold, while pattern-based
modeling has spawned an entirely new approach under the banner of machine
learning that has divided the statistics community (See @Bremman2001,
"The Two Cultures").   Machine learning models can represent almost
arbitrarily complex patterns and incorporate learning and decision making
strategies that have earned them the name 'artificial intelligence.'
Yet at the same time they can be far less transparent their linear
regression predecessors, making both astounding successes and startling
biases and failures like their biological namesake. More ecologists,
particularly theorists and modelers, would do well to learn both how to
take advantage of these strengths and recognize their weaknesses.

@@

Models & Methods
===========================================================================

<!-- Include example of a machine learning prediction that involves biologically
inaccurate predictions such as unbounded growth or growth at zero and yet still
performs well (by fitting well in the region of interest?)  -->

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). "}
```

```{r policies_plot, dependson=c("opt", "plotting-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979]."}
```

```{r sim_plot, dependson=c("sim", "plotting-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."}
```

@@

Discussion
===========================================================================

@@

### 1. Relevant state space

@@

### 2. Predictive accuracy

@@

### 3. Time scale for new data

@@

## Advantages of non-mechanistic models

1. Can better use all of the data available.

@@

2. Better express ambiguity in regions where data is not available

@@

## Better the devil you know?  The dangers of pattern-based approaches

@@

> since the late 1980s there has been some consensus amongst ecologists
that management decisions are best guided by models which are grounded
in ecological theory, and which strike a balance between too much or too
little detail describing the relevant processes (e.g., DeAngelis 1988,
Starfield 1997, Jackson et al. 2000, Carpenter 2003, Nelson et al. 2008)

@@

```{r posteriors, dependson=c("gp", "plotting-options"), fig.cap = "Histogram of posterior distributions for the estimated Gaussian Process shown in Figure 1.  Prior distributions overlaid."}
```

>>>>

\documentclass[author-year, review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
%\linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Beyond Mechanistic Models: The case for machine learning in ecological management},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}
  \title{Beyond Mechanistic Models: The case for machine learning in ecological
         management}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \author[cstar]{Marc Mangel}
  \author[noaa]{Stephan B. Munch}
  \ead{cboettig@ucsc.edu}
  \cortext[cor1]{Corresponding author, cboettig@ucsc.edu}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}
 \end{frontmatter}

\section{Abstract}

\section{Introduction}

@@

While effective machine learning algorithms have speckled the
literature, most theoretical ecologists in particular have been wary of
these `black-box' approaches in favor of simple, mechanistic models.
While a blind application of machine learning approaches would indeed be
dangerous grounds to stake either ecological understanding or management
practice, such concerns are not grounds to reject such methods out of
hand. Rather, further attention is warranted to identify the
circumstances in which machine-learning approaches will most likely
outperform existing more mechanistic approaches, or help crack difficult
management questions where ecological or dynamical complexity has caused
traditional approaches to founder.

@@

Mechanistic models have long been the gold standard of theoretical
modeling in ecology (e.g.~see @Geritz2012 or @Cuddington2013). Only by
understanding the processes involved can we make reliable long term
predictions and build an knowledge of cause and effect that guides the
hypotheses we make, the data we collect, and the management decisions we
make. Process-based models, whether expressed in the language of
mathematics or English, identify the connection between mosquitoes and
the spread of malaria, or greenhouses gases and climate change, guiding
our approach to understand and manage these threats. Despite this
central importance, we argue here that ecologists would do well to give
greater attention to the role non-mechanistic models can play in
ecological management and decision making. The value of these approaches
is greatest in a context where decisions are made over short time
horizons, and updated as new data becomes available.

\subsection{The alternative to process-based models: pattern-based
modeling}

Before we proceed further it would be useful to define our terms and lay
some greater context for the discussion. Throughout, we will distinguish
``mechanistic'' or ``process-based'' models from models that are merely
``correlative'' or ``pattern-based''. Historically pattern-based
modeling meant regression (usually linear regression) and resided in
methodology rooted in statistics departments and was the primary focus
of empirical ecologists, while mechanistic modeling meant dynamical
systems (ODEs, PDEs, SDEs and their discrete kin), residing in
methodology from the mathematics department and was the primary focus of
theoretical ecologists.

We believe these historical divides continue to color much of the
literature today, with theorists more skeptical than empiricists of
statistical methodology and vice versa. Meanwhile, the ground underneath
has shifted. Originally pattern-matching approaches could be critiqued
on the grounds of their simplicity (not everything is linear) while
mechanistic approaches could be critiqued on their tenuous connections
to data -- model parameter values such as death rates, birth rates, etc.
would be estimated in advance and then stuck into the model, rather than
estimated in the context of the model itself. Today both of these
critiques are outdated. Computational power and hierarchical statistical
methods (particularly approximate likelihood or simulation techniques
such as particle filters and approximate Bayesian computing) have
brought the inference richer dynamical systems into their fold, while
pattern-based modeling has spawned an entirely new approach under the
banner of machine learning that has divided the statistics community
(See @Bremman2001, ``The Two Cultures''). Machine learning models can
represent almost arbitrarily complex patterns and incorporate learning
and decision making strategies that have earned them the name
`artificial intelligence.' Yet at the same time they can be far less
transparent their linear regression predecessors, making both astounding
successes and startling biases and failures like their biological
namesake. More ecologists, particularly theorists and modelers, would do
well to learn both how to take advantage of these strengths and
recognize their weaknesses.

@@

\section{Models \& Methods}

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. (The GP is conditioned on
0,0, shown as a pseudo-data point).}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/policies_plot.pdf}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/sim_plot.pdf}
\caption{Gaussian process inference outperforms parametric estimates.
Shown are 100 replicate simulations of the stock dynamics (eq 1) under
the policies derived from each of the estimated models, as well as the
policy based on the exact underlying model.}
\end{figure}

@@

\section{Discussion}

@@

\subsubsection{1. Relevant state space}

@@

\subsubsection{2. Predictive accuracy}

@@

\subsubsection{3. Time scale for new data}

@@

\subsection{Advantages of non-mechanistic models}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Can better use all of the data available.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Better express ambiguity in regions where data is not available
\end{enumerate}

@@

\subsection{Better the devil you know? The dangers of pattern-based
approaches}

@@

\begin{quote}
since the late 1980s there has been some consensus amongst ecologists
that management decisions are best guided by models which are grounded
in ecological theory, and which strike a balance between too much or too
little detail describing the relevant processes (e.g., DeAngelis 1988,
Starfield 1997, Jackson et al. 2000, Carpenter 2003, Nelson et al. 2008)
\end{quote}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/posteriors.pdf}
\caption{Histogram of posterior distributions for the estimated Gaussian
Process shown in Figure 1. Prior distributions overlaid.}
\end{figure}

Reed, William J. 1979. ``Optimal escapement levels in stochastic and
deterministic harvesting models.'' \emph{Journal of Environmental
Economics and Management} 6 (4) (dec): 350--363.
doi:10.1016/0095-0696(79)90014-7.

\end{document}
